How many researchers use 'boilerplate' statistical analysis sections?

### 2021-03-15 manuscript/draft_manuscript.Rmd
update manuscript results plos
update manuscript results

All _PLOS ONE_ articles are freely accessible via the PLOS Application Programming Interface (API). This functionality enabled us to conduct semi-automated searches of full-text articles and analyse data on individual records, including text content and general attributes such as publication date and field(s) of research. To find papers with a statistical methods section we used targeted API searches followed by article filtering based on section headings. The data were downloaded on xx-xxxx-xxx.

@@

Increasing the number of clusters decreases the cluster quality. This indicated that the data comprised of one large cluster and multiple small clusters. Word cloud representing the clustering solutions based on five and ten topics are presented in Figures \@ref(fig:plot-5-topics) and \@ref(fig:plot-10-topics). Common topics reflected statistical software, hypothesis testing by student's t-test and one-way analysis of variance.

Two topics concentrated on the use of Prism/GraphPad (Topic 3, n = 10,940; 9.8%) and SPSS (Topic 5, n = 18,029; 16%) software. A review of records with greatest alignment to Topic 3 showed clear overlaps in text used:

* 10.1371/journal.pone.0043519: All statistical analyses were performed using GraphPad Prism (GraphPad Software Inc., San Diego, CA).
* 10.1371/journal.pone.0045453: GraphPad Prism (Graphpad Software, San Diego, CA) was used for all analyses.
* 10.1371/journal.pone.0170640: All statistical analysis of the data was performed using GraphPad Prism software (GraphPad Prism Software Inc., La Jolla, Ca).
* 10.1371/journal.pone.0036750: All statistical analysis was performed using Graphpad Prism software.
* 10.1371/journal.pone.0058148: All statistical analysis was performed using the Graphpad Prism software.

Boilerplate text to describe statistical methods for between-group comparisons was identified in Topic 1 (n = 12,531; 11%) and Topic 4 (n = 16,740; 15%)
Topic 1 had the smallest number of unique terms in both five (#terms = 54,780) and ten (#terms = 24,686) topic solutions. Statistical methods sections that were strong matches to Topic 1 focused on hypothesis testing using Student's t-test and definitions of statistical significance:

* 10.1371/journal.pone.0077872: Student’s t-test was used for statistical analysis. A p value of <0.05 was considered statistically significant.
* 10.1371/journal.pone.0144950: Student’s t-test was used for statistical analysis. A p value < 0.05 was considered to be statistically significant.
* 10.1371/journal.pone.0019412: The Student t-test was used for all statistical analyses, and a P value less than 0.05 was considered statistically significant.
* 10.1371/journal.pone.0055941: Statistical analysis was performed using a two-tailed Student’s t-test, and p-values <0.05 were considered statistically significant.
* 10.1371/journal.pone.0101993: Statistical analysis was performed using Student's t-test. P-values <0.05 were considered statistically significant.

Top matches for Topic 4 followed a template of summarising data by means with standard errors, formal analysis by one-way analysis of variance and post hoc mulitple comparisons for testing statistical significance:

* 10.1371/journal.pone.0098207: Data are presented as the mean ±SEM. The data were analyzed using the Student's t-test for comparison of two groups, or by one-way analysis of variance (ANOVA) followed by Tukey's post-hoc test for multiple comparisons. Comparisons with p-values less than 0.05 were considered significant.

* 10.1371/journal.pone.0051180: Results are presented as mean ± SEM. Comparison between two groups were analyzed using Student’s t test. Comparisons among multiple groups were performed using one-way analysis of variance (ANOVA), followed by Tukey’s test.

* 10.1371/journal.pone.0207844: Values are presented as mean ± SEM. Differences were analyzed between groups using one-way analysis of variance (ANOVA) followed by a post hoc Tukey’s multiple comparison to determine statistical significance.

* 10.1371/journal.pone.0046301: All data were presented as mean ± SEM. Statistical significance was assessed with one-way analysis of variance (ANOVA) followed by a post hoc (Bonferroni) test for multiple group comparison. Differences with p-value less than 0.05 were considered statistically significant.

* 10.1371/journal.pone.0145237: Results were presented as Mean ±SEM. Mann-Whitney’s U test was used to compare two groups. For multiple group comparisons, one-way ANOVA tests were performed followed up by Tukey’s post-hoc multiple comparisons test. P< 0.05 was considered statistically significant.

Top matches (Topic 5):

* 10.1371/journal.pone.0120046: Statistical analysis was performed using the SPSS version 18.0 (SPSS Inc., Chicago, IL, USA). Mann-Whitney test was used for the analysis of continuous variables. Chi-square test, Fisher’s exact test or one-way ANOVA test was employed for the analysis of categorical variables. P < 0.05 was considered statistically significant.

* 10.1371/journal.pone.0204950: Statistical analyses were performed using SPSS version 20.0 for Windows (SPSS Inc., Chicago, IL). Categorical and continuous variables were evaluated with the non-parametric chi-square test, the Kruskal-Wallis test, and the Mann-Whitney U test. All statistical tests were two-sided, and a p value <0.05 was considered statistically significant.

* 10.1371/journal.pone.0062685: SPSS version 21.0 for Windows (SPSS, Inc., Chicago, IL, USA) was used for statistical analysis. Categorical variables were analyzed by a chi-square test or Fisher’s exact test. Continuous variables were analyzed by independent t-test or Mann-Whitney test. Logistic regression analysis was performed to evaluate the effects of independent variables on clinical outcomes. A p-value of <0.05 by the two-tailed test was considered statistically significant.

* 10.1371/journal.pone.0133783: Data are expressed as means ± SD or medians (with interquartile ranges). Statistical analyses were performed using SPSS software version 19.0 (SPSS, Inc., Chicago, IL, USA). Statistical significance for intergroup differences was assessed by the χ2 test or Fisher’s exact test for categorical variables, and by the Student’s t-test, or Mann–Whitney U test for continuous variables. Correlations between variables were tested with Spearman’s correlation analysis. A P-value <0.05 was considered significant for all tests.

* 10.1371/journal.pone.0098797: Univariate analysis was performed using Mann–Whitney U-tests for continuous variables and Pearson's chi-square or Fisher's exact tests for categorical variables. Statistical significance was set at a P-value <0.05. SPSS statistical software version 19 (SPSS, Chicago, IL, USA) was used to perform the statistical analyses.

>>>>

All _PLOS ONE_ articles are freely accessible via the PLOS Application Programming Interface (API). This functionality enabled us to conduct semi-automated searches of full-text articles and analyse data on individual records, including text content and general attributes such as publication date and field(s) of research. To find papers with a statistical methods section we used targeted API searches followed by article filtering based on section headings. The data were downloaded on 3 July 2020.

@@

Classifying statistical methods sections into two clusters showed a general split between means-based testing and model-based analysis. Increasing the number of clusters decreased cluster quality based on global goodness-of-fit measures (Supplementary Figure - Silhouette + between- & within-cluster dispersions). This indicated that the data comprised one large, heterogeneous cluster and multiple smaller clusters. To examine the effect of increasing the number of clusters on cluster content, we examined themes present among smaller clusters under five-topic (\@ref(fig:plot-5-topics)) and ten-topic (\@ref(fig:plot-10-topics) solutions. Examples of content among top matches for each topic are presented in Table X.

Topic 1 had the smallest number of unique terms under both five (n = 12,531; 54,780 unique terms) and ten (n = 3,775; 24,686 unique terms) topic solutions. Sections assigned to Topic 1 focused on the use of Student's t-test followed by statements of statistical significance. Under the ten topic solution, Topic 1 was split into two clusters based on the use of 5% (Topic 1: n = 3,775) versus multiple thresholds for declaring statistical significance (Topic 9: n = 6,104).

Two topics concentrated on statistical software representing 26% and 17% of all statistical methods sections under five and ten topic solutions. Review of top matching articles assigned to Topic 3 (Prism GraphPad) revealed evidence of boilerplate text, with many omitting details about statistical methods used. For sections citing the use of SPSS, increasing the number of clusters appeared to isolate references to software only versus articles that provided further details on statistical methods used.

Topic 4 was characterised by the use descriptive statistics followed by between-group testing including post-hoc multiple comparisons. Descriptive statistics were commonly reported as means with standard deviations or standard errors. Related topics focused on analysis of experimental designs (Topic 10: n = 4,664) and two-group analyses only (Topic 6: n = 4,746).

Remaining topics reflected applications of meta-analysis (Topic 7: n = 5,165), regression/survival analysis (Topic 2: n = 19,478) and larger sections that combined data processing and analysis (Topic 8: n = 38,183) [TODO: more work needed on this]

and  Larger clusters focused on more advanced statistical methods
